---
title: "002 - The Ames real estate data"
output: html_notebook
---

The Ames housing data set (De Cock 2011) is an excellent resource for learning about models that we will use throughout this book. It contains data on 2,930 properties in Ames, Iowa, including columns related to

* house characteristics (bedrooms, garage, fireplace, pool, porch, etc.),
* location (neighborhood),
* lot information (zoning, shape, size, etc.),
* ratings of condition and quality, and
* sale price.  



```
Our goal for these data is to predict the sale price of a house based on its other characteristics.
```  

The raw data are provided by the authors, but in our analyses in this book, we use a transformed version available in the modeldata package. This version has several changes and improvements to the data7. For example, the longitude and latitude values have been determined for each property. Also, some columns were modified to be more analysis ready. For example:

In the raw data, if a house did not have a particular feature, it was implicitly encoded as missing. For example, there were 2,732 properties that did not have an alleyway. Instead of leaving these as missing, they were relabeled in the transformed version to indicate that no alley was available.

The categorical predictors were converted to R’s factor data type. While both the tidyverse and base R have moved away from importing data as factors by default, this data type is a better approach for storing qualitative data for modeling than simple strings.

We removed a set of quality descriptors for each house since they are more like outcomes than predictors.
To load the data:

```{r}
library( tidyverse )   # We live in the tidyverse.
library( modeldata )   # Ames data is in  this package.
library( cowplot )     # For professional graphics.
library( scales )      # For professional scales.

data( ames )          # Fetch the Ames data.

ames                  # Display portion to show we have the correct data. 
```

4.1 IMPORTANT FEATURES

It  makes sense to start with the outcome we want to predict: the last sale price of the house (in USD):

```{r}
ames %>% 
  ggplot( aes( x = Sale_Price )) +
    geom_histogram( bins = 25,
                    color = "black",
                    fill  = "green" )
```
I prefer to use a density histogram and too draw a smoothed  density plot with the 
histogram.  I also use the ggplot theme theme_bw().  In addition when dealing with dollar amounts greater that are in the thousands of dollars it is useful to change the unit to thousands of dollars ($k) rather than dollars($)
```{r}
                                
ames %>% 
  mutate( Sale_Price = Sale_Price / 1000) %>% 
  ggplot( aes( x= Sale_Price, y = ..density..)) +
    geom_histogram( bins = 25,
                    color = "black",
                    fill  = "green" ) +
    geom_density( color = "black",
                  linetype = "dashed" ) +
    xlab( "Sale Price(Thousand dollars)") +
    ylab( "Density" ) +
    ggtitle("Ames residential real estate prices") +
    theme_cowplot()
  

      

```
  
  The data are right-skewed; there are more inexpensive houses than expensive ones. The median sale price was $160,000 and the most expensive house was $755,000. When modeling this outcome, a strong argument can be made that the price should be log-transformed. The advantages of doing this are that no houses would be predicted with negative sale prices and that errors in predicting expensive houses will not have an undue influence on the model. Also, from a statistical perspective, a logarithmic transform may also stabilize the variance in a way that makes inference more legitimate. Let’s visualize the transformed data:

```{r}
ggplot(ames, aes(x = Sale_Price)) + 
  geom_histogram(bins = 25,
                 color = "black",
                 fill = "green") +
  scale_x_log10()
```

Again we will change this by working in units of thousands of dollars.

```{r}
ames %>% 
  mutate( Sale_Price = Sale_Price/1000 ) %>% 
    ggplot( aes( x= Sale_Price )) +
      geom_histogram( bins = 25,
                      color = "black",
                      fill = "green" ) +
      scale_x_log10() + 
      expand_limits( x = 1500 ) + 
      xlab( "Sale price(Thousand dollars)") +
      ylab( "Density" ) +
      ggtitle( "Log10 transform os Sale price" ) +
      theme_cowplot()

```

This is a much more symmetrical distribution although it appears that we we probably have an outlier in the lower tail.  So it appears that the log transform is useful.

```{r}
minsp <-min( ames$Sale_Price)
maxsp <- max(ames$Sale_Price)
r     <- maxsp/minsp
fminsp <- format( minsp, digits = 0, big.mark = "," )
fmaxsp <- format( maxsp, digits = 0, big.mark = "," )format
fr     <- format( r, digits = 2 )
```


Incidentally, there is the the law of ten, which says that if the ratio of the maximum values divided by the minimum value the logarithm transformation will be useful in obtaining a symmetric distribution. (I don't currently have the full reference for this but I know it is in the book **Applied Regression Analysis** by Sanford Weisberg.)  

In the Ames data the maximum sales price is $755,000 and the minimum sale price is $12,789.  The ratio is 59.04 which is greater than 10 so it is worth while trying the log transformation.

Therefore we are going to update the Ames data set set by replacing sale price by the logarithm of the sale price.  Since we are using United states dollars we will use logs to the base 10.  

We do the transformation below.  We will save resulting data file in the data directory for our project as as RDS file which is more efficient.  Our R code is:

```{r}
################################################################################
# We store are transformed response variable data in ames1 so we don't overwrite
# our orginal data in x=case we need it.  I have seen many individuals lose
# their data by overwriting their orginal data.
################################################################################
ames01 <-                                      
  ames %>% 
    mutate( Sale_Price = log10( Sale_Price ))

##################################################################################
# Verify the transformation was made by looking at the first 10 sale prices
##################################################################################

ames01$Sale_Price[1:10]

```
We now save the data so we do not need to recreate it every time we do more analysis.  We save the data using R-RDS files which are very efficient to write and read.

We also store our data files in a sub-directory.  If you keep your program or scripts in the same directory as your data it too easy to delete something by mistake.

In the code chunk below we save the files:  


```{r}
################################################################################
# Save the data as an RDS file.
################################################################################

dd <- file.path( getwd(), "data" )          # Build data directory

fp <- file.path( dd, "ames.RDS")            # To save original data.

saveRDS( ames, fp )                         # Save orginal data 

fp <- file.path( dd, "ames01.RDS")          # Path to transformed data.

saveRDS( ames01, fp  )                      # Save logged Sale Price file.

dir( path = dd )                            # Verify files saved.

```

Another important aspect of these data for our modeling are their geographic locations. This spatial information is contained in the data in two ways: a qualitative Neighborhood label as well as quantitative longitude and latitude data.

The most important predictor variable of real estate price is location.  We have two different types of predictor variables that deal with location.  On is the house's neighborhood. This is a qualitative variables.  Typically, within the neighborhood there will be a price trends that reflect the age of houses and the desirability.  

For example I live in Arlington Texas.  The neighborhood is adjacent to Lake Arlington.  The streets closer to lake Arlington tend to have higher prices than the streets that are more distant from the lake.  

Dr. Slinkman's remark.  The book we are using has maps of the neighborhood is Ames Iowa.  As so as I get time I will add them to bu solution.  If you know how to draw statistical maps I would be most grateful if you would add them to this project.  



